---
title: "Effect of Police on Ethnic Protests"
author: "Jazmia Henry"
header-includes:
    - \usepackage{setspace}\doublespacing
date: "May 10, 2018"
output: pdf_document
---

#Introduction              
While the First Amendment is normally evoked in common parlance to denote the right to free speech, other four protections under the Constitutional amendment are important as well. One right in particular, the right to assemble, has been used by marginalized members in society to advocate for equal rights and protections under the law. In 2009, Stanford Professor Susan Olzak created a dataset that combined every ethnic protest in the United States from 1954 through 1995. These protests were ones initiated by oppressed groups against governmental institutions. This dataset was collected so that researchers could study and compare the events of ethnic protests from 1954-1995. Instead of using the data to answer questions about collective action, I will be answering a research question of my own:"Does the presence of police officer escalate or de-escalate violence that happens during or directly after and as a result of ethnic protests?"            
            
#Methodology                      
To answer my overarching research question, I begin by converting the ICPSR data from factor variables to numeric variables. I do this by creating a labels function by sorting the levels of my variables and separating these levels into the amount of possible responses of each variable. I get the number of possible responses by referring to the data's codebook. I then create the recoded variable by running the as.numeric function against the number of possible responses in my variable and the now stripped original variable as coded in the dataset. Lastly, I combine the created label variable and the stripped original variable to create a new numeric variable. I found this way of recoding factor variables to numeric variables from ICPSR. *See .RMD for code.*              
The variables that I used in this paper are:             
\begin{itemize}        
\item "size": a variable that collects the size of the protests from 0 which stands for unknown size to 5 which is over a thousand members.      
\item "reg": which is a a variable coded 1-4 for region of the country where the protest took place. 1 is for the Northest, 2 is for "North Central" (What we consider the Midwest), 3 is for South, and 4 is for West.       
\item "race1": a variable that codes the race of the initiating group. 0 is for unknown race, 1 is for African-American or Black, 2 is for White and Non-Hispanic, 3 is for Hispanic which Includes Mexicans, Puerto Ricans, Latin Americans, and Filipinos, 4 is for Asian people Includes Japanese, Chinese, Southeast Asians, Taiwanese, people from Hong Kong, and 5 is for Mixed racial identification.         
\item "race2-5": three variables that codes the races of that join a protest to assist on behalf of race1. The codes are the same as race1.        
\item "thrt": a variable that records the threat of violence in a particular protest. Code 0 stands for no threat of violence, 1 is threat present as either (a) personal warning, or (b) physical evidence of weapons mentioned in report.Code 2 is random or undirected violence such as sniping, firebombs, people throwing bricks, or gunshots heard. Code 3 is violence which is aimed at specific individuals or properties such as specific stores looted and or damaged or vandalism to a specific ethnic organization or house.        
\item "prop": a variable that records the amount of property damage done as a result of the protest. 0 means no damage, 1 is very little damage to property like a broken window, 2 is damage to buildings, cars, homes, or other structures, and 3 is extensive damage like fire-bombing or looting reported.      
\item "take": a variable that records if a space or building has been taken over or occupied by protesters. Code 0 means no takeover, 1 means Barricades used or blocking access to entry by protesters, 2 is a Takover of space including whole streets, vacant lots, 3 means Occupying buildings, and 4 is Taking hostages or preventing someone from leaving the building.            
\item "weap": a variable coding the type of weapons used by protesters. Code 0 is no weapons reported at the event, 1 is non-lethal weapons such as rocks, bricks, bottles thrown, clubs, or mace, and 2 is lethal weapons used like guns, knives, firebombs.          
\item "pol": is a binomial variable that codes whether police where present to contain the crowd of protesters. 0 is police were not present and 1 is police were present.             
\item "pollest": an estimate of the amount of police forces present. Code 0  means no police were present, 1 means uncertainty, 2 means a Small number of police present (1-10), 3 is medium number of police present (11-50) 4 is Medium-Large number of police present (51-100), 5 is large number of police present (Over a hundred police), and 6 is massive forces present (500+).          
\item "pollev": records the type of police present. Code 0 means police not present, 1 is local police present only, 2 is Local police plus reinforcements from some other source, such as reserves, or police from neighborhood precincts, 3 is local and state police, 4 is local police and National Guard, 5 is local Police and Army, and 6 is Army and National Guard.             
\item "pollacts": a variable that records the most intrusive acts done by the police. Code 0 is No police present, 1 is Police present but no actions taken, 2 is police actions are limited and unobtrusive and no physical force used, 3 is police used at least some physical force  such as forming lines, or grappling
with protestors, carrying them from the site, making arrests, or using their nightsticks, 4 is police used weapons such as nightsticks, guns, teargas, firehoses, or cattleprods.            
\end{itemize}           
Within this paper, I will be using instrumental variable regression analysis to tests my theory and hypothesis. My theory is that police presence in ethnic protests actually escalate violence and whether or not the police is present has more to do with the race of the protesters than the weapons the protesters use or whether protesters have vandalized or occupied property. I hypothesize that:     
**1. There is extreme racial bias in the patrolling of ethnic protests, and that police presence at ethnic rallies leads to more violence.**             
**2. When weapons are used or buildings are occupied with white protesters, even if the police is present, police is less like to use physical force.**             
**3. When other races of protesters assist marginalized races in protests, then this mitigates the amount of force that the police will use against protesters as well as the type of police force present.**                      
**4. The type of police force present is indicative of police escalation so the Army and National Guard will be more likely to use physical force than local police alone.**               
```{r introduction, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include= FALSE, warning= FALSE)
library(prettyR)
load("34341-0001-Data.rda")
EP<-da34341.0001
lbls <- sort(levels(EP$size))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$size <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$size))
EP$size <- add.value.labels(EP$size, lbls)
lbls <- sort(levels(EP$reg))
lbls <- (sub("^\\([0-4]+\\) +(.+$)", "\\1", lbls))
EP$reg <- as.numeric(sub("^\\(0*([0-4]+)\\).+$", "\\1", EP$reg))
EP$reg <- add.value.labels(EP$reg, lbls)
lbls <- sort(levels(EP$race1))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$race1 <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$race1))
EP$race1 <- add.value.labels(EP$race1, lbls)
lbls <- sort(levels(EP$thrt))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$thrt <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$thrt))
EP$thrt <- add.value.labels(EP$thrt, lbls)
lbls <- sort(levels(EP$prop))
lbls <- (sub("^\\([0-3]+\\) +(.+$)", "\\1", lbls))
EP$prop <- as.numeric(sub("^\\(0*([0-3]+)\\).+$", "\\1", EP$prop))
EP$prop <- add.value.labels(EP$prop, lbls)
lbls <- sort(levels(EP$take))
lbls <- (sub("^\\([0-4]+\\) +(.+$)", "\\1", lbls))
EP$take <- as.numeric(sub("^\\(0*([0-4]+)\\).+$", "\\1", EP$take))
EP$take <- add.value.labels(EP$take, lbls)
lbls <- sort(levels(EP$weap))
lbls <- (sub("^\\([0-2]+\\) +(.+$)", "\\1", lbls))
EP$weap <- as.numeric(sub("^\\(0*([0-2]+)\\).+$", "\\1", EP$weap))
EP$weap <- add.value.labels(EP$weap, lbls)
lbls <- sort(levels(EP$pol))
lbls <- (sub("^\\([0-1]+\\) +(.+$)", "\\1", lbls))
EP$pol <- as.numeric(sub("^\\(0*([0-1]+)\\).+$", "\\1", EP$pol))
EP$pol <- add.value.labels(EP$pol, lbls)
lbls <- sort(levels(EP$pollev))
lbls <- (sub("^\\([0-6]+\\) +(.+$)", "\\1", lbls))
EP$pollev <- as.numeric(sub("^\\(0*([0-6]+)\\).+$", "\\1", EP$pollev))
EP$pollev <- add.value.labels(EP$pollev, lbls)
lbls <- sort(levels(EP$polest))
lbls <- (sub("^\\([0-6]+\\) +(.+$)", "\\1", lbls))
EP$polest <- as.numeric(sub("^\\(0*([0-6]+)\\).+$", "\\1", EP$polest))
EP$polest <- add.value.labels(EP$polest, lbls)
lbls <- sort(levels(EP$polacts))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$polacts <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$polacts))
EP$polacts <- add.value.labels(EP$polacts, lbls)
lbls <- sort(levels(EP$race2))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$race2 <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$race2))
EP$race2 <- add.value.labels(EP$race2, lbls)
lbls <- sort(levels(EP$race3))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$race3 <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$race3))
EP$race3 <- add.value.labels(EP$race3, lbls)
lbls <- sort(levels(EP$race4))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$race4 <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$race4))
EP$race4 <- add.value.labels(EP$race4, lbls)
lbls <- sort(levels(EP$race5))
lbls <- (sub("^\\([0-5]+\\) +(.+$)", "\\1", lbls))
EP$race5 <- as.numeric(sub("^\\(0*([0-5]+)\\).+$", "\\1", EP$race5))
EP$race5 <- add.value.labels(EP$race5, lbls)
```
#Results                   
I begin my analysis by running instrumental variables regression of the pol viariable against the threat and size variable. I do this to see if there is a relationship between the presence of police at a protest and the threat being posed by the protesters and their size. Instrumental variable regression "J.iv1" shows that there is a positive correlation between the threat that a group of protesters make and the presence of police. As well there is an 11% increase in the likelihood of police being present for every one unit increase in the size of the protesting crowd.                   
When I add the race of the initiating group of protesters to my instrumental variable analysis in model "J.iv2", size is no longer an important predictor. Here, the most important predictor is race, which makes me feel confident about my first hypothesis. In J.iv2, I use race1, region, and size as instruments that affect my model. By adding race1 as a predictor, I get interesting results. First, "thrt" is no longer a positive predictor of police presence. In fact, a naive reading of this analysis suggests that the more threat that a group of protesters pose to the peace, the less likely the police is to be present. Instead, I think that this points to my theory and second hypothesis: the threat of physical violence that protesters pose to the public does not cause the police to patrol the protest. Police patrols a protest according to the race of the people involved in the protest.              
In my next regression analysis "J.iv3", I get interesting results: race 1 and thrt are still important predictors and race two, even when interacting against race 1, is not seen as an important predictor. I extend my analysis in "J.iv4" which uses the same predictors but regresses them against "polacts" to test my third hypothesis. My analysis shows that race and threats are important predictors, but, just as my J.iv2 and J.iv3 regressions, thrt is a negative predictor. Substantively, this analysis seems to suggest that threats made by protesters will inspire police presence and perhaps even unintrusive police force such as keeping them contained to one area. On the other hand, if the protesters are black, then they are likely to be contained to one area and may even be handled violently. (Being around a 2.5 intercept suggests equal likelihood of a "2" response- police containment- or a "3" response- physical force.)                
To test my last hypothesis, I run "J.iv5" to see if police levels inspires police officers to behave more aggressively. Here, I disprove my hypothesis and find that threat level and the race of the protesters is the most important predictor of violent police action. Due to this result, I am going to run a similar regression, this time with police level as the outcome variable and race, threat, and size as a way of seeing if the introduction of outside police forces can be predicted by the race of protestors as well.This analysis shows that the threat posed by protesters is an extremely significant negative predictor of if outside police forces will be involved. Black protesters are more likely to have outside police forces called against them than any other race. If the intiating protestor is white, Asian, Mixed Race, or Hispanic, then they are less likely to get the threat of outside police forces coming against them. What is interesting here is that the presence of other races leads to an increase in the possibility of outside police forces. This means that if a White, Hispanic, Mixed Race, or Asian group of protesters assist with another initiating race, then they are more likely to have police forces come against them.          
               
##An Examination of Initiating Race                                                   
When separating black fom white protesters and dropping the rest of the initiating races, I am able to truly test the validity of my second hypothesis while at the same time looking closer at the race1 variable.               
In my first instrumental variable regression analysis "J.iv7", I find that while being black increases the possibility of police presence, the "thrt" and "size" variables are more statistically significant predictors when white and black is turned into a dichotomous variable. What is a predictor of police being present, then, is not simple if black people are protesting, but if a large number of black people are protesting or if they are seen as being a threat to the public. The question would be if the people are seen as threatening because they are black, if the perceived threat of provocation is heightened when black people are the actors, or if there is an assumption of violent behavior when the protestors are black, but I have no way of testing this with this dataset.               
When running black against polacts in "J.iv8", I get the same thing. Being black has a positive relationship with violent police action, but it does not have a statistically significant relationship. The threat that the crowd poses is, once again, the most significant predictor of the police behaving violently. Interestingly, the size of the group does not matter here which makes me continue to wonder if the perception of threats is heightened when the actors are black.               
In my last regression analysis using my recoded "black" variable, my analysis shows that blackness is the most important predictor of outside police forces as well as the size. Substantively, this means that when there is a large gathering of black protesters, there is a 12% increase in the possibility that the local police will be joined by an outside police force. When black protestors are present, then there is an 18% increase in the probability that the local police will enlist help from an outside police force.            
My last regression analysis looks at my second hypothesis directly.This one proposes that even when weapons are used and buildings are occupied, when the protesters are white, then the police is less likely to behave violently. My analysis in "J.iv10" is inconclusive. The p-values show that none of the predictors are statistically significant.            


```{r Results, results=TRUE, echo=FALSE}
library(AER)
library(stargazer)
white<-recode(EP$race1, "2=1; 1=0; else=NA")
black<-recode(EP$race1, "1=1; 2=0; else=NA")
j.iv1<-ivreg(pol~thrt+size, data=EP)
summary(j.iv1)
j.iv2<-ivreg(pol~thrt+race1+size|race1+size+reg, data=EP)
summary(j.iv2, diagnostics=TRUE)
j.iv3<-ivreg(pol~thrt+race1*race2+race3+size|race1+size+reg, data=EP)
summary(j.iv3, diagnostics=TRUE)
j.iv4<-ivreg(polacts~thrt+race1+race2+race3+size|race1+size+reg, data=EP)
summary(j.iv4, diagnostics=TRUE)
j.iv5<-ivreg(polacts~pollev+thrt+race1+race2+race3+size|race1+size+reg, data=EP)
summary(j.iv5, diagnostics=TRUE)
j.iv6<-ivreg(pollev~thrt+race1+race2+race3+size|race1+size+reg, data=EP)
summary(j.iv6, diagnostics=TRUE)

j.iv7<-ivreg(pol~black+thrt+size|black+size+reg, data=EP)
summary(j.iv7, diagnostics=TRUE)
j.iv8<-ivreg(polacts~black+thrt+size|black+size+reg, data=EP)
summary(j.iv8, diagnostics=TRUE)
j.iv9<-ivreg(pollev~black+thrt+size|black+size+reg, data=EP)
summary(j.iv9, diagnostics=TRUE)
j.iv10<-ivreg(polacts~white+weap+prop+thrt+size|white+size+reg, data=EP)
summary(j.iv10, diagnostics=TRUE)
```
              
#Conclusion         
In conclusion, I find that the right to assemble is a right that has traditionally been racially defined. Black protesters have not enjoyed that right as much as other races. Though my dataset only includes the years 1954-1995, I would not be surprised if race would still be a predictor of treatment, though it would, hopefully, not be as strong of a predictor. The rhetoric being espoused by our current leaders does not make me hopeful.        
While the data is very valuable, I found myself restricted by many of the definitions of the variables. The "thrt" variable in particular may be racially biased, for reports of perceived threats may be increased when the protesters are black. Though the "thrt" variable was created by researchers depending on newspaper articles and police reports, I know that historically, these devices have been inaccurate when reporting the behavior of black people (see the testimony of Carolyn Bryant whom lied about Emmett Till's conduct and got him killed). 
If I had more time, I would have fixed the analysis that I tried in my "R tryout" chunk that can be found in the .RMD file. Unfortunately, the year has come to an end much quicker than I anticipated so I ran out of time before I could fix the problems in my code. 
```{r Try out, echo=FALSE, include=FALSE, results=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning= FALSE, include=FALSE)
p<-1579 #police present
np<-1171 #no police present
police<-rep(0, 2750)
nopolice<-rep(0, 2750)
pff<-779+180 #police physical force (pff used to distinguish from pp)
pff
p.v<-p - pff #number of times police behaved violently
p.v
npff<-1579-p.v
npff #no police violence
police<-c(police, rep(1, p.v))
nopolice<-c(nopolice, rep(1, p.v))
noviolence<-c(rep(1, npff), rep(0, pff))
.3927*p #percent of instances ending in physical violence when police present
#620 times police present ended in violence, but police violence not the only violence. 
 #code 2 and 3 of "EP$thrts" combined to examine threats by protesters
thrt.npv<-1195-620 #threat by protesters minus police responding violently
nopv<-3945-thrt.npv #number of observation of threats plus the number of instances to get a total number of observations introduced in dataset
police<-c(police, rep(1, thrt.npv))
nopolice<-c(nopolice, rep(1, thrt.npv))
noviolence<-c(rep(1, thrt.npv), rep(0, nopv))
#attempted to turn into a datset but was unable to due to uneven rows. I, unforttunately, did not have time to ask for assistance with this, but wanted to show my attempts in good faith. Plus, I spent many hours working on this chunk alone and wanted to show what I tried. This would have been a good addition to my paper and analysis if I were able to figure it out in time.
```


